Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North

### 2023-04-14 analysis/paper/paper.qmd
progress

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
institute:
  - wlu:
      name: Wilfrid Laurier University
      # address: 23 Science Street, Eureka, Mississippi, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template_modified.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed in an heterogeneous sampling area straddling two of Northern Ontario’s ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We've recently developed an open-source workflow for camera trap image identification, access, and storage, which we are using to identify images, with next steps involving the development of an image classifier.

@@

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/WILDLab/rofcamtrap/')
backup_options <- options()
options(scipen = 1, digits = 2)
library(magrittr)
```

@@

The James Bay Lowlands in Northern Ontario hold some of the world’s last intact wild places but it is under pressure from resource extraction and global warming. Our understanding of basic biodiversity measures like species occurrence, abundance, diversity, and density for the region are still rudimental but needed to weigh ecological-economical decision tradeoffs.   
Traditionally, quantifying and monitoring biodiversity in northern regions required expensive and time-consuming field surveys. But today, the use of non-intrusive methods, such as arrays of 100s of camera traps, has significantly changed field practices for reliably quantifying wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification, and processing, of image datasets.

Machine learning methods, such as deep neural networks, are now in use to automate the process of species identification and help in that endeavour. We aim to evaluate the viability of deep neural networks for camera trap surveys currently taking place in Northern Ontario’s heterogeneous landscapes, including the Hudson Bay Lowlands and Ontario Shield ecozones, as variables like weather, seasonality, vegetation type, and species characteristics can impact the performance of these methods. There are currently no available machine learning models for identifying species from images in northern regions and developing one would be beneficial as a monitoring and modeling tool for northern boreal ecosystems.

The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed in an heterogeneous sampling area straddling two of Northern Ontario’s ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We've recently developed an open-source workflow for camera trap image identification, access, and storage, which we are using to identify images, with next steps involving the development of an image classifier.

@@

### Camera deployment

The camera deployment protocol follows a spatially balanced hierarchical design that provides an even distribution of sample sites across biotic and abiotic conditions within the study area. The study sites span three ecoregions and two ecozones. The network was established by CWS in 2020 and seasonally equipped in a rolling design with 500 autonomous recording units (ARUs) to monitor audible species diversity paired with site-specific on-ground vegetation imagery. In 2022, 232 unbaited wildlife camera traps (CTs) were added to a subset of study sites to quantify mammalian occurrence and diversity. @fig-deployment shows the location of the and their current status. retrieved cameras were retrieved in September 2022.

@@

cams_sf <- RoF_CameraDeploymentLog_WLU %>% 
  dplyr::filter(!is.na(latitude)) %>% 
  dplyr::filter(!is.na(longitude)) %>% 
  sf::st_as_sf(coords=c("longitude","latitude"), crs="EPSG:4326") %>% 
  dplyr::mutate(is_retrieved = ifelse(!is.na(date_retrieved), "Yes", "No"))

@@

```{r}
#| label: fig-deployment
#| echo: false
#| fig-cap: "Camera deployment status"

@@

```{r}
#| label: fig-deployment-poster
#| eval: false

p <- base +
  # ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOREGION),inherit.aes = FALSE,
  #                     size = 2.5, data = regions_labels) +
  ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords %>% 
                        dplyr::filter(is_retrieved == "Yes"), cex = 1, shape=21, 
                      col="black", fill = "red",
                      ggplot2::aes(x=X, y=Y), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill = "Camera \nRetrieved")
p
```

@@

Images were retrieved from all cameras using a SD card reader and copied onto a 10 TB hard drive of brand LaCie. The images amounted to 1.8TB of memory on disk, for a total of __`r nb_image_fmt`__ images. Images were stored with the following directory structure:

- Plot id (e.g. P128)
- Site id (e.g. P128-1)
- Camera id (e.g. CFS-10)
- DCIM folder (created by the camera)
- Overflow folder (e.g. 100RECNX, each such folder contains up to 10,000 images)

The images were processed through the animal detection model MegaDetector, version 5.0, using a Lenovo X1 laptop with a NVIDIA GeForce GTX 1650 Max-Q graphics card, which took about 2 weeks of quasi-continuous processing to complete. The model processes each image and produces “detections” (bounding boxes) at a  variable confidence score.

@@

About __`r dets_tb_percent[1]` %__ of the images, or __`r dets_tb_fmt[1]`__ images, were found by the model to possibly have either a person, animal, or vehicle. The model did not produce detections for __`r dets_tb_fmt[2]`__ images, or __`r dets_tb_percent[2]` %__ of the total.

@@

This resulted in a total of __`r dets_tb_cat_unfil_sum_fmt`__ detections. The totals for each category are: __`r dets_tb_cat_unfil_fmt[1]`__ animals (or __`r dets_tb_cat_unfil_percent[1]` %__); __`r dets_tb_cat_unfil_fmt[2]`__ persons (or __`r dets_tb_cat_unfil_percent[2]` %__); and __`r dets_tb_cat_unfil_fmt[3]`__ vehicles (or __`r dets_tb_cat_unfil_percent[3]` %__).

Using a cut-off value for confidence threshold of 0.1, this left us with a total of __`r dets_tb_cat_sum_fmt`__ detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). The totals for each category are: __`r dets_tb_cat_fmt[1]`__ animals (or __`r dets_tb_cat_percent[1]` %__); __`r dets_tb_cat_fmt[2]`__ persons (or __`r dets_tb_cat_percent[2]` %__); and __`r dets_tb_cat_fmt[3]`__ vehicles (or __`r dets_tb_cat_percent[3]` %__).

```{r}
#| label: fig-detections-histogram
#| echo: false
#| fig-cap: "Histogram of detection confidence scores."
#| fig.width: 8
#| fig.height: 4.5

# Histogram of confidence scores above 0.1
dets_0.1 %>% 
  dplyr::mutate(category_id = as.character(category_id)) %>% 
  dplyr::mutate(Type = dplyr::case_match(category_id,
                                         "0" ~ "Empty",
                                         "1" ~ "Animal",
                                         "2" ~ "Person",
                                         "3" ~ "Vehicle")) %>% 
  ggplot2::ggplot() +
  ggplot2::theme_bw() +
  ggplot2::geom_histogram(ggplot2::aes(x=confidence, group=Type, fill=Type),
                          bins = 100) +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::xlab("Confidence Score") +
  ggplot2::ylab("Count")
```

Finally, it is important to note that not only pictures with detections were uploaded to the platform. Because camera traps produce images in bursts, we included all the pictures from a given bursts if any of the images in the bursts include a detection with certainty above the threshold.

@@

- *Deployable online*. This was to allow for for multiple taggers
- *Flexible tagging interface*. Few platforms allow for a full customization of the tagging interface, down to the smallest details.
- *Allowing to tag multiple bounding boxes per image*. Most platforms only allow for tagging the entire image, and not for portions of it (e.g. bounding boxes of detections produced by MegaDetector). This is necessary for when we will crop our images to train our neural network model. 
- *Optionally, open source,* to allow for maximum reproducibility of the process.

We looked at platforms not traditionally used for tagging camera trap images, but used by the larger community of machine learning practitioners for which the task of tagging is a common step, often referred to as labeling. We found two platforms that matched all those criteria: LabelStudio and CVAT. After testing both platforms, LabelStudio was chosen as it was simpler to implement. The Community edition of LabelStudio was deployed on an Ubuntu virtual machine hosted by Compute Canada, using an Apache server and a NGINX reverse proxy, which is a standard approach for custom hosting of websites.

@@

@@

Out of the __`r dets_tb_cat_fmt[1]`__ animal detections, __`r processed_fmt`__ have been processed on the platform (or __`r processed_pct` %__) for a total of __`r processed_img_fmt`__ images (i.e. __`r processed_img_pct` %__ of the total amount of images, or __`r processed_img_up_pct` %__ of the total amount of images uploaded to the platform). This does not include the empty images also viewed as part of a set, bringing that amount to 92,826 “detections” and __`r viewed_fmt`__, (i.e. __`r viewed_pct` %__ of the total amount of images, or __`r viewed_pct_up` %__ of the total amount of images uploaded to the platform).

@@

Out of the __`r processed_fmt`__ detections, __`r dels_nb_fmt`__ were false positives (or __`r dels_nb_pct` %__), and __`r non_mans_nb_fmt`__ were true positives (or __`r non_mans_nb_pct` %__), to which __`r mans_nb_fmt`__ false negatives were added (manually added bounding boxes, __`r mans_nb_pct` %__). Out of the __`r non_mans_nb_fmt`__ true positives, __`r orig_tb_fmt[3]`__ had to have their bounding box adjusted,

<!-- ### Issues encountered

The main issue we encountered was a problem in the performances of MegaDetector which caused a lot of false positives, as the model was often tricked by vegetation. -->

@@

We analysed the results of the tagging process under two main forms: presence absence data for each camera; and relative species abundance measured as the number of days, for a given month, that a given species was detected by a given camera.

@@

```{r}
sum(unlist(lapply(extr_list, function(x) sum(x$cats == "NA"))))
sum(unlist(lapply(extr_list, function(x) is.na(x$cats))))
#  [1] "1"   "2"   "8"   "9"   "12"  "15"  "16"  "17"  "18"  "11"  "13"  "14"  "19"  "20"  "23"  "157" "247" "21"  "22"

pairs(extr_table_sub_tb, col = "dodgerblue")
round(cor(extr_table_sub_tb), 2)
```

## Preliminary results

### Community data

Using the number of days detected as a measure of relative abundance, the most common species (detected more than 10 days in the season) are listed in Table 1 (the full table is available in the appendix at the end of this report). Note the large amount detections in the "bird" and "duck" species categories which could be further distinguished into specific species.

@@

sp_tab %>%
  dplyr::filter(days_detected >= 10) %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Relative abundances for species above 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
  ) %>% 
  gt::opt_stylize(style = 6, color = "gray")

@@

As for the spatial distribution of those relative abundances, @fig-map-pies show relative species abundances across the landscape, for species detected more than 50 days in the season (note that the here the detections have grouped by plot number for better visibility, with a plot containing up to 4 cameras) . There are no clear spatial patterns, but some tendencies can be noted: Moose detections are predominant in the northwestern of the sampling array, while caribou and sandhill cranes detections share the most of the detections for the rest of the array.

```{r}
#| label: fig-map-pies
#| echo: false
#| fig-cap: "Relative species abundances at plot sites (only species species detected more than 50 days in the season are shown)"
#| fig.width: 8
#| fig.height: 6

@@

base +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long")
```

@@

@fig-rda-pa shows the triplot for the RDA on presence-absence data, which yielded a low $R^2$ of `r vegan::RsquareAdj(pa_rda)[2]`, but which is still relevant for exploring species associations. Land cover might not be able to explain community structure, but it can still show which species are found together and around what kind of land cover class. Here, we see that caribou and sandhill cranes are associated with open and treed wetlands (bogs and fens) while black bear and moose are found together in more forested areas. Canada geese are, as is to be expected, associated with water while smaller species like squirrels and martens score away from the larger mammals in a mix of habitats. Note that a full legend of the land cover abbreviations are to be found in the appendix of this report.

```{r}
#| label: fig-rda-pa
#| echo: false
#| fig-cap: "Triplot of RDA on presence-absence data"

custom_rda_plot(pa_rda)
```
```{r}
# Jaccard pcoa

>>>>

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Samantha McFarlane:
      institute: eccc
  - Jennifer Baltzer:
      institute: wlu
  - Cheryl A. Johnson:   
      institute: eccc
  - Mathieu Leblond:
      institute: eccc
  - Eric Neilson:
      institute: nrcan
  - Philip Weibe:
      institute: nrcan
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
  - Josie Hughes:
      institute: eccc
institute:
  - wlu:
      name: Wilfrid Laurier University
  - eccc:
      name: Environment & Climate Change Canada
  - nrcan:
      name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template_modified.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/WILDLab/rofcamtrap/')
backup_options <- options()
options(scipen = 1, digits = 12)
library(magrittr)
```

@@

Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and global warming. Our understanding of basic biodiversity measures like species occurrence, abundance, diversity, and density for the region are still rudimental but needed to weigh ecological-economical decision tradeoffs. 
Traditionally, quantifying and monitoring biodiversity in northern regions required expensive and time-consuming field surveys. Today, the use of non-intrusive methods, such as arrays of 100s of camera traps, has significantly changed field practices for reliably quantifying wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification, and processing, of image datasets.

Machine learning methods, such as deep neural networks, are now in use to automate the process of species identification and help in that endeavor. We aim to evaluate the viability of deep neural networks for camera trap surveys currently taking place in Northern Ontario’s heterogeneous landscapes, including the Hudson Bay Lowlands and Ontario Shield ecozones, as variables like weather, seasonality, vegetation type, and species characteristics can impact the performance of these methods. The extent to which these methods perform for northern regions is not well established. There are currently no available machine learning models for identifying species from images in northern regions and developing one would be beneficial as a monitoring and modeling tool for northern boreal ecosystems.

Using an array of cameras deployed in an heterogeneous sampling area straddling two of Northern Ontario’s ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage, which we are using to identify images, with next steps involving the development of an image classifier.

@@

### Camera deployment and setup

The camera deployment protocol follows a spatially balanced hierarchical design that provides an even distribution of sample sites across biotic and abiotic conditions within the study area (@vanDamBates2018). The study sites span three ecoregions (Northern Taiga, James Bay, Big Trout Lake) and two ecozones (Hudson Bay Lowlands, Ontario Shield). The network was established by CWS in 2020 and seasonally equipped in a rolling design with 500 autonomous recording units (ARUs) to monitor audible species diversity paired with site-specific on-ground vegetation imagery. In March through June of 2022 193 unbaited and unlured wildlife camera traps ([Reconyx Hyper Fire II](https://reconyx.com/product/hyperfire-2-covert-ir-camera)) were added to this study design. All cameras were all programmed to take bursts of 5 images (one per second) at each trigger of their motion sensor.

Briefly, 5 ARUs are deployed per site, with 100 m between ARUs. Up to 4 cameras were paired with ARUs at each of these sites, with at least one of these cameras placed at a random ARU site and one other prioritized to focus on a game trail or open area to increase detection probability (in total, we have 43 plots with 4 cameras, 5 plots with 3 cameras and 3 plots with 2 cameras). @fig-deployment shows the location of the clustered cameras and their current retrieval status. 170 cameras were retrieved in September 2022, totaling a deployment spanning 5-7 months. 5 additional cameras were only retrieved in March 2023 due to logistical and inaccessibility issues. There remains 18 cameras in the field to be retrieved in 2023.

The work conducted under this project focuses on quantifying and developing an open-source workflow to quantifying the wildlife species detected in those images that have been retrieved to date.

@@

cams_sf <- RoF_CameraDeploymentLog_WLU %>% 
  dplyr::filter(!is.na(latitude)) %>% 
  dplyr::filter(!is.na(longitude)) %>% 
  sf::st_as_sf(coords=c("longitude","latitude"), crs="EPSG:4326", remove=F) %>% 
  dplyr::mutate(is_retrieved = ifelse(!is.na(date_retrieved), "Yes", "No"))

@@

```{r}
#| label: fig-deployment
#| echo: false
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."

@@

```{r}
#| label: fig-deployment-poster
#| eval: false
#| echo: false

# base <- ggmap::ggmap(map) +
#   ggplot2::labs(x = "Longitude", y = "Latitude") +
#   # ggplot2::geom_polygon(data = pols, ggplot2::aes(x=X, y=Y), col = 1, fill = NA)
#   ggplot2::geom_path(data = pols, ggplot2::aes(x=X, y=Y), col = 1, linewidth=0.2)
# 
# p <- base +
#   # ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOREGION),inherit.aes = FALSE,
#   #                     size = 2.5, data = regions_labels) +
#   ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
#                       size = 2.5, data = regions_labels) +
#   ggplot2::geom_point(data = cams_sf_coords %>% 
#                         dplyr::filter(is_retrieved == "Yes"), cex = 1, shape=21, 
#                       col="black", fill = "red",
#                       ggplot2::aes(x=X, y=Y), 
#                       inherit.aes = FALSE) +
#   ggplot2::labs(fill = "Camera \nRetrieved")
# p
```

@@

Images were retrieved from all cameras using a 32 GB SD card and copied onto a 10 TB hard drive (LaCie d2 Professional). The images amounted to 1.8TB of memory on this disk, for a total of __`r nb_image_fmt`__ images. Images were stored with the following directory structure:

- Plot id (e.g. P128)
- Site id (e.g. P128-1)
- Camera id (e.g. CFS-10, with the first 3 letters corresponding to the group owning the camera: CFS for Canadian Forest Service, WLU for Wilfrid Laurier University)
- DCIM folder (created by the camera)
- Overflow folder (e.g. 100RECNX, 101RECNX ; each such folder contains up to 10,000 images)

The images were processed through the animal detection model MegaDetector (@beery2019efficient), version 5.0, using a Lenovo X1 laptop with a NVIDIA GeForce GTX 1650 Max-Q graphics card, which took about 2 weeks of quasi-continuous processing to complete. The model processes each image and produces “detections” (bounding boxes) at a  variable confidence score.

@@

At this first step into the process, no filtering was applied beyond the model predictions which classifies images as either likely to contain an object of relevance or empty. About __`r dets_tb_percent[1]` %__ of the images, (__`r dets_tb_fmt[1]`__) images, were found by the model to possibly contain either a person, animal, or vehicle. The model did not produce detections for __`r dets_tb_fmt[2]`__ images, (__`r dets_tb_percent[2]` %__) of the total.

@@

Because the model can produce more than one detection per image, this resulted in a total of __`r dets_tb_cat_unfil_sum_fmt`__ detections. The totals for each category are: __`r dets_tb_cat_unfil_fmt[1]`__ animals (or __`r dets_tb_cat_unfil_percent[1]` %__); __`r dets_tb_cat_unfil_fmt[2]`__ persons (or __`r dets_tb_cat_unfil_percent[2]` %__); and __`r dets_tb_cat_unfil_fmt[3]`__ vehicles (or __`r dets_tb_cat_unfil_percent[3]` %__).

Megadetector is an AI model which produces confidence scores with each of its predictive detections. With no filtering, the overwhelming majority of the detections are of very low confidence (85% of detections are below a score of 0.1). This explains the high number of "person" and "vehicle" detections which are of course unrealistic. It is recommended to filter low probability detections, and cut-off values of 0.1 or 0.2 are usually suggested in the Megadetector documentation.

Using a cut-off value for the detection confidence of 0.1 (,i.e. discarding all detections with confidence below 0.1), this left us with a total of __`r dets_tb_cat_sum_fmt`__ detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). After applying this confidence threshold (decided on by the Wildlife Working Group) the total images to process  for each category are:  __`r dets_tb_cat_fmt[1]`__ animals (__`r dets_tb_cat_percent[1]` %__); __`r dets_tb_cat_fmt[2]`__ persons (__`r dets_tb_cat_percent[2]` %__); and __`r dets_tb_cat_fmt[3]`__ vehicles (__`r dets_tb_cat_percent[3]` %__).

```{r}
#| label: fig-detections-histogram
#| echo: false
#| fig-cap: "Histogram of detection confidence scores produced for each bounding box predicted by Megadetector, showing a bimodal distribution. The colors represent the three categories of object identified by Megadetector: animals, persons and vehicles. Note that the hsitogram is truncated at our selected confidence threshold of 0.1."
#| fig.width: 8
#| fig.height: 4.5

# Histogram of confidence scores above 0.1
dets_0.1 %>% 
  dplyr::mutate(category_id = as.character(category_id)) %>% 
  dplyr::mutate(Type = dplyr::case_match(category_id,
                                         "0" ~ "Empty",
                                         "1" ~ "Animal",
                                         "2" ~ "Person",
                                         "3" ~ "Vehicle")) %>% 
  ggplot2::ggplot() +
  ggplot2::theme_bw() +
  ggplot2::geom_histogram(ggplot2::aes(x=confidence, group=Type, fill=Type),
                          bins = 100) +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::xlab("Confidence Score") +
  ggplot2::ylab("Count") +
  ggplot2::geom_vline(ggplot2::aes(xintercept = 0.1), lty=2, col = "firebrick")
```

Finally, it is important to note that not only pictures with detections were uploaded to the LabelStudio platform. As stated before, camera traps produce images in bursts (in our case, 5 images per trigger). It is possible for Megadetector to miss a detection within a burst, or to produce variable confidence scores within that burst, which could lead to missing true detections if not all pictures from a given burst are filtered in. Therefore, we included all the images from a given burst if any of the images in the burst include a detection with certainty above the threshold, including potentially empty images from that burst.

@@

- *Deployable online*. This was to allow for multiple people to help process the images.
- *Flexible tagging interface*. Few platforms allow for a full customization of the tagging interface, down to the smallest details.
- *Ability to tag multiple bounding boxes per image*. Most platforms only allow for tagging the entire image, and not for portions of it (e.g. bounding boxes of detections produced by MegaDetector). This is necessary for when we will crop our images to train our neural network model. 
- *Optionally, open source,* to allow for maximum reproducibility of the process.

To guide our choice, we used these criteria to informally compare 51 different platforms and tools currently available for processing and managing camera trap images. We first looked at platforms traditionally used for tagging camera trap images, but found them often lacking in two main aspects: flexibility of the tagging interface, and ability to manipulate bounding box information.

These two features are often better addressed by tools not traditionally used for tagging camera trap images, but used by the larger community of machine learning practitioners. Indeed, machine learning datasets outside can be very diverse and often more detailed than camera trap datasets, requiring processing tools with more detailed features. Tagging is a common step in machine learning research and in industry, and if often referred to as labeling.

We found two platforms that matched all of our criteria: LabelStudio (@LabelStudio) and CVAT (@boris_sekachev_2020_4009388). After testing both platforms, LabelStudio was chosen because the data format it uses is more easily compatible with Megadetector's COCO format output, making it easier to setup. We deployed the Community edition LabelStudio (version 1.6.0rc5) on an Ubuntu virtual machine hosted by Compute Canada, using an Apache server and a NGINX reverse proxy, which is a standard approach for custom hosting of websites.

@@

saveRDS(anns_wide_noempty, "analysis/data/derived_data/annotations_wide_noempty.rds")

@@

Out of the __`r dets_tb_cat_fmt[1]`__ animal detections, __`r processed_fmt`__ have been processed to date using our platform (or __`r processed_pct` %__) for a total of __`r processed_img_fmt`__ images (i.e. __`r processed_img_pct` %__ of the total amount of images, or __`r processed_img_up_pct` %__ of the total amount of images uploaded to the platform).

As stated above, all images of a given burst are viewed if at least one image in the burst contains a detections with confidence above the threshold. This include potentially empty images. If we add the empty images viewed as part of a burst, it brings the total amount to __`r viewed_fmt`__ images, (i.e. __`r viewed_pct` %__ of the total amount of images, or __`r viewed_pct_up` %__ of the total amount of images uploaded to the platform).

@@

Out of the __`r processed_fmt`__ detections, __`r dels_nb_fmt`__ were false positives (or __`r dels_nb_pct` %__), where Megadetector identified something, but there was nothing. In addition, __`r non_mans_nb_fmt`__ were true positives (or __`r non_mans_nb_pct` %__), to which __`r mans_nb_fmt`__ false negatives were added (manually added bounding boxes, __`r mans_nb_pct` %__). Out of the __`r non_mans_nb_fmt`__ true positives, __`r orig_tb_fmt[3]`__ had to have their bounding box adjusted. The below summarises information for the tagging progress.

```{r}
#| label: tab-summary-table
#| echo: false

sum_tab <- tibble::tribble(
  ~Status,                ~Images,           ~Detections,
  "Total",                nb_image_fmt,      dets_tb_cat_unfil_sum_fmt,
  "With detections",      dets_tb_fmt[1],    "N/A",
  "Empty",                dets_tb_fmt[2],    "N/A",
  "Filtered (> 0.1)",     uploaded_fmt,      dets_tb_cat_sum_fmt,
  "Processed",            processed_img_fmt, processed_fmt
)

knitr::kable(sum_tab,
             caption = "Summary table on tagging progress")
```

The relatively high rate of false positives is attributable to a combination of factors. First, the high seasonality of Ontario's far north makes it so that it is difficult to predict whether a camera that was set up in the winter won't be overgrown by vegetation in the spring. Locations where vegetation has grown in front of the sensor, combined with windy events, creates a perfect storm for false cameras triggers when vegetation is moving in front of the sensors. In addition, Megadetector is a likely to mistake a certain proportion of those vegetation pictures for actual animals. We are currently exploring avenues to tackle this roadblock by identifying repeating bounding images and by using a model to sort false from true positives.

#### Reproducibility

We are committed to producing a reproducible workflow. All code necessary to reproduce this report is available at an online [repository](https://github.com/StewartWILDlab/rofcamtrap). The two software tools used in our workflow are also available at online repositories:

- [**mdtools**](https://github.com/StewartWILDlab/mdtools): a python-based command line tool for converting Megadetector outputs into LabelStudio inputs, and LabelStudio outputs into CSV tables.

- [**mdscripts**](https://github.com/StewartWILDlab/mdscripts): a bash-based command line tool for running Megadetector and mdtools.

@@

We summarized our existing species data to date for analysis in two main ways:   (1) species presence/absence data for each camera and, (2) relative species abundance, measured as the number of days, within each month that each  species was detected at a camera site.

@@

```{r}
#| label: extract-cats
#| echo: false
#| eval: false

# sum(unlist(lapply(extr_list, function(x) sum(x$cats == "NA"))))
# sum(unlist(lapply(extr_list, function(x) is.na(x$cats))))
# #  [1] "1"   "2"   "8"   "9"   "12"  "15"  "16"  "17"  "18"  "11"  "13"  "14"  "19"  "20"  "23"  "157" "247" "21"  "22"
# 
# pairs(extr_table_sub_tb, col = "dodgerblue")
# round(cor(extr_table_sub_tb), 2)
```

## Data exploration and preliminary results

### Wildlife community compposition

We obtained repeat detections of at least 31 mammal and bird species across 193 sites and 5-7 months of detections The most common species detected (detected more than 10 days in the season) are listed in Table 1 (the full table is available in the appendix at the end of this report). Note the large amount detections in the “bird” and “duck” species categories which could be further distinguished into specific species.

@@

# sci_names <- taxize::comm2sci(sp_tab$species)
# saveRDS(sci_names, "analysis/data/derived_data/sci_names.rds")
sci_names <- readRDS("analysis/data/derived_data/sci_names.rds") %>% 
  unlist() %>% tibble::as_tibble(rownames = NA) %>% 
  tibble::rownames_to_column() %>% 
  dplyr::rename(Scientific_name = value,
                species = rowname) %>% 
  tibble::add_row(Scientific_name = "Ursus americanus", species="Black bear") %>% 
  tibble::add_row(Scientific_name = "Perisoreus canadensis", species="Canada Jay")

sp_tab %>%
  dplyr::left_join(sci_names, by="species") %>% 
  dplyr::relocate(Scientific_name, .after=species) %>% 
  dplyr::filter(days_detected >= 10) %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Relative abundances for species with more than 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
    Scientific_name = "Scientific name"
  ) %>% 
  gt::tab_style(style = list(
      gt::cell_text(style="italic")),
    locations = gt::cells_body(columns = Scientific_name)) %>% 
  gt::opt_stylize(style = 6, color = "gray")

@@

As for the spatial distribution of those relative abundances, @fig-map-pies shows relative species abundances across the landscape, for species detected more than 50 days in the season (this number is roughly equal to the standard deviation of the distribution of detection days, which has a mean of 32 days). Note that for this figure, the detections have grouped by plot number for better visibility, with a plot containing up to 4 cameras (see deployment methods). There are no clear spatial patterns, but some generalities can be noted: Moose detections are predominant in the northwestern and southeastern zones of the sampling array, while caribou and sandhill crane detections share the most of the detections for the central part of the array.

```{r}
#| label: fig-map-pies
#| echo: false
#| fig-cap: "Pie chart map of relative species abundances across our study area Detections have grouped by plot for better visibility, with a plot containing up to 4 cameras. Only species detected more than 50 days in the season are shown for clarity."
#| fig.width: 8
#| fig.height: 6

@@

regions_labels_bis <- zones_filter %>% 
  sf::st_drop_geometry() %>% 
  dplyr::bind_cols(data.frame(x = c(-85, -88),
                              y = c(52, 52.6)))

base +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long") +
    ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels_bis)
```

@@

To start exploring general species-habitat associations we completed an redundancy analysis (RDA, using the `vegan` R package, @vegan). RDAs allow to study the relationships between two matrices of data, and is often used in community studies, using a constraining matrix of environmental variables for each site and a corresponding matrix of species observations (either presence absence or relative abundances). Ordinations also combine very well with clustering analyses to produce insights about associations between sites.
